Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-05 paper.tex
deleted the alternative proof of L2 that was commented out
to end of Prop3

    % Alternative: Induction:
    %
    % Induction on $k = d(T,R)$:
    %
    % Basis($k = 1$):
    % If rank of $\parent(n+1)$ changes between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = 0$ and $|\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 0 \leq 1 - 1 = d(T,R) - |\rank_T(\parent(n+1)) - \rank_R(\parent(n+1))|$.
    % If on the other hand the rank of $\parent(n+1}$ does not change between $T$ and $R$, it is $d(T{\big|}_n, R{\big|}_n) = d(T,R) = 1$ and $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = 1$ and therefore, $d(T{\big|}_n, R{\big|}_n) = 1 \leq 1 - 0 = d(T,R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    % Step($k \to k+1$):
    % Let $T$ and $R$ be trees on $n+1$ taxa with distance $d(T,R) = k$.
    % Let $R'$ be the last tree on a shortest path from $T$ to $R$, i.e. $d(T,R') = k$.
    % We can apply the induction hypothesis on $T$ and $R'$ and see that $d(T{\big|}_n, R'{\big|}_n) \leq d(T, R') - |r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| := k - m$.
    % Now we differ two cases:
    % \begin{enumerate}
    %     \item $\parent(n+1}$ changes its rank between $R$ and $R'$ or is affected by the $\nni$ move is performed between $R$ and $R'$.
    %
    %     Then the trees $R{\big|}_n$ and $R'{\big|}_n$ are equal. \todo{Is this obivous?}
    %     Therefore, it is $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m$.
    %     It is either $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m+1$, $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m$, or $|r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})| = m-1$.
    %     In either case, we have $d(T{\big|}_n,R{\big|}_n) = d(T{\big|}_n, R'{\big|}_n) \leq k - m = (k+1) - (m+1) \leq d(T, R) - |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    %
    %     \item $\parent(n+1}$ has same rank in $R$ and $R'$.
    %
    %     In this case it is $d(T{\big|}_n,R{\big|}_n) \leq d(T{\big|}_n,R'{\big|}_n) + d(R'{\big|}_n, R{\big|}_n) \leq d(T{\big|}_n, R'{\big|}_n) + 1 \leq (k-m)+ 1$.
    %     It also follows that $|r_{T}(\parent(n+1}) - r_{R'}(\parent(n+1})| = |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$ and therefore, we can conclude $d(T{\big|}_n,R{\big|}_n) \leq k-m+1 = k+1-m = d(T,R) -  |r_{T}(\parent(n+1}) - r_{R}(\parent(n+1})|$.
    % \end{enumerate}

%Following Proposition might not be needed(?)

Related to Lemma~\ref{lemma:distance_delete_taxon} is Proposition~\ref{proposition:lower_bound_distance} which provides a lower bound for the distance of two trees.
Using the same arguments as in the proof of Lemma~\ref{lemma:distance_delete_taxon}, it easily follows that the rank difference of parents of the same taxon but in two different trees gives such a bound.

\begin{proposition}
    Given trees $T$ and $R$ with taxon $x$ that fulfils $|\rank_T(\parent(x)) - \rank_R(\parent(x))| = k$, it follows $d(T,R) \geq k$.
    \label{proposition:lower_bound_distance}
\end{proposition}

@@

Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
\todo{Do we need to explain Bubble Sort?}

>>>>

Using the same arguemtn as in the proof of Lemma~\ref{lemma:distance_delete_taxon} we can establish the following Proposition~\ref{proposition:lower_bound_distance} that provides a lower bound on the distance between trees.

\begin{proposition}
If trees $T$ and $R$ and taxon $x$ are such that $|\rank_T(\parent_T(x)) - \rank_R(\parent_R(x))| = \delta$ then $d(T,R) \geq \delta$.
\label{proposition:lower_bound_distance}
\end{proposition}

@@

Following the observations above, only swaps of the first three elements of a sequence with Bubble Sort need to be adapted for the computation of shortest caterpillar paths.
Let us consider the first three elements $a,b,c$, appearing in this order in a sequence representing a caterpillar tree on a caterpillar path between trees $T$ and $R$.
\todo{Check throughout that $\parent$ is never used without tree name in the index.}
If the rank of $\parent(c)$ is higher than the ranks of $\parent(a)$ and $\parent(b)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $a$ and $b$ is on the other side of $c$ in $R$, this taxon swaps position with $c$ in the sequence.
And if both $a$ and $b$ are on the other side of $c$ in $R$, the taxon that has the parent with highest rank of all these three taxa in $R$ swaps with $c$ at first.
This makes sure that there is no additional swap of $a$ and $b$ needed on the subsequent path to $R$.
\todo{Do we need to explain Bubble Sort?}
